---
title: "Advanced Functions in R"
page-navigation: false
format:
  html: 
    toc: false
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

Welcome to the module on Advanced Functions in R. This module introduces the core programming tools that make R a truly flexible and powerful language. You’ll learn how to automate repetitive tasks, design your own reusable functions, and connect multiple operations together into clear, efficient workflows. Each tutorial section will conclude with a short knowledge quiz.

1. [Pipes](/tutorials/pipes/index.qmd)
2. [Loops](/tutorials/loop/index.qmd)
3. [User-Defined Functions](/tutorials/user-fun/index.qmd)

Each tutorial will comprise a series of interactive lessons with practice problems embedded throughout. Each tutorial section will conclude with a short knowledge quiz.

## Learning Goals

After completing this module, you will be able to:

### Pipes

1. Understand how the `%>%` and `|>` pipe operators work.
2. Build readable, multi-step pipelines that chain together data transformations.
3. Use placeholders and the exposition pipe `%$%` for more flexible function composition.

### Loops

1. Use `for` loops to iterate across elements of vectors or data frames.
2. Write conditional loops (`while` and `repeat`) and control flow with `break` and `next`.
3. Recognize when vectorized operations are a more efficient alternative to looping.

### User-defined Functions

1. Write your own functions to automate repetitive tasks and simplify code.
2. Use arguments and default values effectively.
3. Apply conditional logic and return values to control function output.

Click the button below to get started with the first tutorial.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/pipes/index.html")
```
