---
title: "More on User Defined Functions"
---

### Multiple Arguments and Default Values

Functions in R can take multiple arguments, and some of these can have default values. This means that a function can work with just the essential inputs, while still allowing flexibility when you want to adjust its behavior.

```{r}
power <- function(x, exp = 2) {
  x ^ exp
}

power(3)

power(3, 3)
```

In this example, if you don’t specify `exp`, it defaults to 2, so the function squares its input by default.

You can also include conditional statements such as `if` inside a function, allowing it to behave differently depending on the input.

```{r}
categorize_age <- function(age) {
  if (age < 18) {
    "minor"
  } else {
    "adult"
  }
}

categorize_age(17)
```

This simple example shows how a function can make logical decisions. Conditional logic is especially useful when your function needs to classify, filter, or adjust its behavior dynamically based on its inputs.

A function can return any kind of R object — not just a single value. It can output a vector, a list, a data frame, or even a plot. For example:
```{r}
summarize_vector <- function(vec) {
  mean_val <- mean(vec, na.rm = TRUE)
  sd_val <- sd(vec, na.rm = TRUE)
  n <- sum(!is.na(vec))
  
  return(list(mean = mean_val, sd = sd_val, n = n))
}


summarize_vector(c(1, 2, 3, 4, NA))
```

Here, the function returns a list containing the mean, standard deviation, and number of non-missing values from the input vector.

When you define a function, R automatically creates a new environment for it to run in. Any variables created inside the function exist only within that environment — they are not available outside the function once it finishes running.

```{r, error=TRUE}
add_one <- function(x) {
  y <- x + 1
  return(y)
}

add_one(4)

y
```

In this example, `y` exists only while the function is being executed. Once the function finishes, the temporary variable disappears. This isolation prevents functions from accidentally modifying your global workspace.

Functions can still access variables defined outside them (in the global environment), but they won’t modify those variables unless you explicitly return a value and reassign it.

```{r}

x <- 10
increment <- function() {
  x + 1
}
increment()

x
```

This separation between the function’s internal environment and the global workspace is a fundamental principle of R. It ensures that functions are self-contained, predictable, and safe to reuse without causing unwanted side effects.

## Anonymous Functions and Shortcuts

Sometimes, you might need a simple one-off function to perform a quick operation — for example, as part of another function call. In these cases, defining a named function can feel unnecessary. These short, temporary functions are known as anonymous functions.

You can define them inline using the regular `function()` syntax:

```{r}
sapply(1:5, function(x) x^2)
```

Anonymous functions are especially useful inside tidyverse pipelines when you need to apply a transformation directly within a data workflow:

```{r}
library(dplyr)
data <- data.frame(age = c(20, 25, 30, 35, 40),
                   weight = c(150, 160, 170, 180, 190),
                   height = c(65, 66, 67, 68, 69))
data|> mutate(age_sq = sapply(age, function(x) x^2))
```

They let you write small, focused operations inline — keeping your code concise without cluttering your workspace with extra function names.

::: practice

## Practice

a. Write a function called that takes a single numeric argument and returns a character string describing whether the number is `"positive"`, `"negative"`, or `"zero"`. 

b. Add a second argument to your function that allows the user to specify whether they want the output in uppercase or lowercase letters. The default should be lowercase.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.hint >}}

To go from lowercase to uppercase, you can use the `toupper()` function in R.

## {{< var tab_title.solution >}}

```{r}
# a 
describe_number <- function(num) {
  if (num > 0) {
    return("positive")
  } else if (num < 0) {
    return("negative")
  } else {
    return("zero")
  }
}

# b
describe_number <- function(num, to_upper = FALSE) {
  if (num > 0) {
    result <- "positive"
  } else if (num < 0) {
    result <- "negative"
  } else {
    result <- "zero"
  }
  if (to_upper) {
    result <- toupper(result)
  }
  return(result)
}
```
::::

:::


