---
title: "Conditional Loops"
---

Not all loops in R have a fixed number of repetitions. Sometimes you want your code to keep running until a certain condition is met. In these cases, you can use conditional loops, which rely on logical tests rather than predefined sequences. R provides two main forms of conditional looping: the `while` loop and the `repeat` loop.

A `while` loop checks a condition before each iteration and keeps running as long as that condition is TRUE. For example, the following loop prints numbers from 1 to 5:

```{r}
x <- 1
while (x <= 5) {
  print(x)
  x <- x + 1
}

```

Here, the variable `x` starts at 1. Before every iteration, R checks whether `x <= 5`. If the condition is true, the code inside the braces runs. Once the condition becomes false, the loop stops automatically.

A `repeat` loop, on the other hand, does not check a condition at the start—it runs indefinitely until you explicitly tell it to stop using `break()`.

```{r}
x <- 1
repeat {
  print(x)
  x <- x + 1
  if (x > 5) break
}
```

This produces the same output as the previous example. The `repeat` loop is useful when the stopping condition depends on a computation that happens inside the loop, rather than something known in advance.

You can also control the flow inside loops using two special commands: `break` and `next`. `break`, as we saw, immediately stops the loop, even if the condition is still true, while `next` skips the rest of the current iteration and moves on to the next one.

Here’s a simple demonstration:

```{r}
for (i in 1:5) {
  if (i == 3) next     # skip 3
  if (i == 5) break    # stop before printing 5
  print(i)
}

```

This prints 1, 2, and 4 — it skips 3 because of `next`, and stops entirely before reaching 5 because of `break`.

Conditional loops and flow control give you flexibility: you can make loops respond to data dynamically, handle exceptions, or stop automatically when a specific goal is reached.

::: practice

## Practice

Write a conditional loop that print all the possible combination between the numbers 1 to 10 in where the first number is bigger than the second number

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```


## {{< var tab_title.solution >}}

```{r}
i <- 1
while(i <= 10){
  j <- 1
  while(j <= 10){
    if(i < j) break
    print(paste(i, j))
    j <- j + 1
  }
  i <- i + 1
}

```
::::

:::








