---
title: "Using Pipes for Data Manipulation in R"
---

```{r, echo=FALSE}
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


### Building a Workflow with Pipes

Pipes are especially powerful when used to chain together multiple data processing steps. Let’s look at a more substantial example.

Suppose we want to:

1. Center the variable `age` on 18.
2. Create scale scores for `extraversion` and `neuroticism`.
3. Keep only participants aged 18 or older.
4. Select only the scale scores and demographic variables.
5. Sort the data by `extraversion` in ascending order.

Using `dplyr` pipes, we can write:

```{r}
tmp1 <- bfi %>%
  mutate(age = age - 18,
         extra = rowMeans(across(matches("^e\\d$")), na.rm = TRUE),
         neuro = rowMeans(across(matches("^n\\d$")), na.rm = TRUE)) %>%
  filter(age >= 0) %>%
  select(extra, neuro, age, gen, ed, gm) %>%
  arrange(extra)

head(tmp1, 20)
```

Each step is explicit, readable, and self-contained — it’s immediately clear what happens and in what order.

The same logic written in base R would be much harder to follow:

```{r}
tmp2 <- bfi
tmp2$age   <- tmp2$age - 18
tmp2$extra <- rowMeans(tmp2[grep("^e\\d$", colnames(tmp2))], na.rm = TRUE)
tmp2$neuro <- rowMeans(tmp2[grep("^n\\d$", colnames(tmp2))], na.rm = TRUE)
tmp2 <- tmp2[tmp2$age >= 0, c("extra", "neuro", "age", "gen", "ed", "gm")]
tmp2 <- tmp2[order(tmp2$extra), ]
```

Both approaches produce the same result — but the piped version is far clearer.

### How Pipes Work

The pipe operator simply takes the result from the left-hand side and inserts it as the first argument in the function on the right-hand side.

This means the two lines below are equivalent:

```{r}
mean(bfi$age)
bfi$age %>% mean()
```

The same applies here:

```{r}
var(bfi$a1, na.rm = TRUE)
bfi$a1 %>% var(na.rm = TRUE)
```

### What happens when the data isn’t the first argument?

Some functions — like `lm()` — don’t take the data frame as the first argument. If you try to use a pipe directly, it won’t work:

```{r, error=TRUE}
bfi %>% lm(extra ~ age)  # Error
```


In these cases, you can use the special placeholder `.` to tell R explicitly where to insert the piped object:

```{r}
bfi %>% lm(extra ~ age, data = .)
```

This small trick makes it possible to use pipes with nearly any function, even when the data aren’t the first argument.

```{webr-r}
#| autorun: true
#| context: setup
library(psychTools)
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


::: practice

## Practice

Use pipes to perform the following data manipulations on the `bfi` dataset:

1. Create a new variable `age_centered` that centers `age` around its mean.
2. Filter the dataset to include only participants with `age` greater than 25.
3. Select only the new centered age variable, the scores for agreeableness and conscientiousness, and the demographic variables `gen` and `ed`.
4. Display the first 20 rows of the resulting dataset.


:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```


## {{< var tab_title.solution >}}

```{r}
bfi %>%
  mutate(age_centered = age - mean(age, na.rm = TRUE)) %>%
  filter(age > 25) %>%
  select(age_centered, agree, consc, gen, ed) %>%
  head(20)
```
::::

:::


