---
title: "Why Use Pipes in R?"
---

```{r, echo=F}
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


When you perform several transformations in R, your code can quickly become cluttered with nested parentheses or a series of temporary variables. Pipes offer a cleaner, more intuitive alternative: they let you write code that mirrors the logical flow of your analysis. Each step passes its output directly into the next, creating a readable sequence of operations that tells the story of your data transformation from start to finish.

First of all, let's call the the `dplyr` package to use the pipe operator `%>%` and look at the dataset we will work with:

```{r}
library(dplyr)
head(bfi)
```


For example, consider this chain of operations applied to the `bfi` dataset:

1. Select five personality scale scores.
2. Compute their covariance matrix.
3. Extract the variances from this matrix.
4. Take the square roots to obtain standard deviations.

In base R, this can be written as nested function calls:

```{r}
sqrt(
  diag(
    cov(
      bfi[c("agree", "consc", "extra", "neuro", "open")],
      use = "pairwise"
    )
  )
)
```

While technically correct, the structure is hard to follow. Alternatively, we can save intermediate results:

```{r}
tmp <- bfi[c("agree", "consc", "extra", "neuro", "open")]
tmp <- cov(tmp, use = "pairwise")
tmp <- diag(tmp)
sqrt(tmp)
```

This approach is clearer but involves multiple temporary variables and assignments.

Using a pipe, we can express the same sequence in a single, logical flow that reads from top to bottom:

```{r}
bfi %>%
  select(agree, consc, extra, neuro, open) %>%
  cov(use = "pairwise") %>%
  diag() %>%
  sqrt()
```

Here, the result of each function is passed directly into the next, eliminating unnecessary assignments and parentheses. Pipes enhance readability by allowing you to see the sequence of transformations at a glance, making your code easier to understand and maintain.

```{webr-r}
#| autorun: true
#| context: setup
library(psychTools)
data("bfi", package = "psychTools")
library(dplyr) # load dplyr package
library(psych) # load psych package
names(bfi.keys) <- c("agree", "consc", "extra", "neuro", "open")
tmp <- scoreItems(keys = bfi.keys, items = bfi, impute = "none")
bfi <- data.frame(bfi, as.data.frame(tmp$scores)) %>%
     mutate(
        gender = factor(gender, labels = c("male", "female")),
        agree_z = (agree - mean(agree, na.rm = TRUE)) / sd(agree, na.rm = TRUE),
        consc_z = (consc - mean(consc, na.rm = TRUE)) / sd(consc, na.rm = TRUE),
        extra_z = (extra - mean(extra, na.rm = TRUE)) / sd(extra, na.rm = TRUE),
        neuro_z = (neuro - mean(neuro, na.rm = TRUE)) / sd(neuro, na.rm = TRUE),
        open_z  = (open  - mean(open,  na.rm = TRUE)) / sd(open,  na.rm = TRUE),
        gm = case_when(
                  age < 18 & gender == "male"   ~ "boy",
                  age < 18 & gender == "female" ~ "girl",
                  age >= 18 & gender == "male"  ~ "man",
                  age >= 18 & gender == "female"~ "woman",
                  TRUE ~ NA_character_),
        gm = as.factor(gm),
        ) %>%
        rename(ed = education,
               gen = gender) %>%
        rename_with(.fn = tolower, .cols = matches("\\d$"))
```


::: practice

## Practice

Use a pipeline to calculate the square root of the mean of the agreeableness scale score for males in the `bfi` data.

:::: {.panel-tabset}
## {{< var tab_title.editor >}}

```{webr-r}

```

## {{< var tab_title.hint >}}

You can use the `unlist()` function to covert a list to a vector.

## {{< var tab_title.solution >}}

```{r}
bfi %>%
  filter(gen == "male") %>%
  select(agree) %>%
  unlist() %>%
  mean(na.rm = TRUE) %>%
  sqrt()

```
::::

:::

